\chapter{Conclusions}

This thesis represents a comprehensive exploration of position and orientation estimation methods. Within the scope of this work, it was possible to gather extensive information about sensor fusion, data filtration, and signal processing. The thesis introduces a novel approach to utilizing the knowledge of the system structure. The present work opens the topic of using solutions taken from aviation and other technical sciences in the field of robotic manipulators and other ground-based mechanisms or machines. In particular, it is important to highlight the pioneering use of constraints correction in this field of science and the introduction of improvements to the existing methods.\\

In the thesis, theoretical considerations are complemented by the physical implementation. A working system was successfully designed and deployed based on the knowledge gathered. The developed system is a device that employs various sensors and seamlessly cooperates with application software. The system will be used for further research. For this reason, its operation has been further checked, and its correctness has been validated.\\

Certainly, the work does not exhaust the presented topic. In the future, there are several possible directions for developing and enhancing the system. The measuring system will be continuously tested when exploited during the research project experiments. In order to improve estimation results, a hardware revision can be introduced, which would consist in employing high-precision sensors. Through continuous examination, the software will also benefit from the detection of bugs and corner cases.\\


\section*{Acknowledgment}

This work has been supported by the National Science Centre, Poland under grant no. 2022/45/B/ST8/00661.


% This thesis enhances robotic positioning systems by integrating inertial navigation systems, inspired by aviation principles, with knowledge of mechanical design. The work done is part of a project being carried out at the Division of Theory of Machines and Robots. The primary goal is to fill a crucial gap in the existing setup – the absence of position and orientation sensors.\\

% The thesis explores fundamental navigation sensors, recognizing their limitations. It discusses strategies like filtering and sensor fusion to mitigate errors and obtain accurate position and orientation estimates. The proposed solution involves designing the experimental platform. This system modifies inertial navigation to estimate coordinates efficiently as the platform moves, addressing specific motion constraints.\\

% Anticipated outcomes include a deep understanding of position and orientation measurement systems and the design of a prototype tailored for robotic applications. This thesis may improve positioning systems, benefiting both the division's projects and the broader field of robotics.